---
title: "WK8"
output: pdf_document
date: "2025-08-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(purrr)
library(readr)
library(ggthemes)
library(ggeffects)
library(lme4)
library(dplyr)
library(ggplot2)
#install.packages("emmeans")
library(emmeans)      
#install.packages("broom.mixed")
library(broom.mixed)  
library(viridis)      
library(nlme)         
library(mgcv)         
#setwd("C:/Users/Tobyz/Desktop/S.tort-light-growth/Data")
```

*import plant data*

```{r}
plant <- read.csv("Data/WL2-2023_Size_Combined.csv") %>%
  clean_names() %>%
  mutate(survey_date = as.Date(survey_date, format = "%m/%d/%Y"))
summary(plant)
```

*consolidate light measurement to a weekly measurement*

```{r}
#import light data
light_raw <- read_csv("Data/IntBioHalfHourTable_clean.txt")

#weekly measurement
weekly_light <- light_raw %>%
  mutate(
    timestamp = ymd_hms(TIMESTAMP),
    SlrW_Avg = as.numeric(SlrW_Avg),  # turn into number format
    week = floor_date(timestamp, "week")
  ) %>%
  group_by(week) %>%
  summarise(
    weekly_avg_SlrW = mean(SlrW_Avg, na.rm = TRUE),
    .groups = "drop"
  )

# result
print(weekly_light)
```

*Investigate or filter out plants that show negative growth*
#Q: How could we deal with bad observations?
#Solution: find out tolerance value and then filter out observance data lager than the tolerance value
#Figure A: Histogram showing the frequency of decreases in plant height between consecutive surveys. Most negative growth values are close to zero, likely reflecting measurement noise, while a small number of extreme decreases (<=–5 cm) suggest errors and were removed from subsequent analyses.

```{r}
#PID
plant_growth <- plant %>%
  unite("PID", genotype:rep, sep = "_", remove = FALSE) %>%
  mutate(survey_date = as.Date(survey_date))

#find out plants with negative growth
plant_growth %>%
  arrange(PID, survey_date) %>%  # arrange in time sequence
  group_by(PID) %>%
  mutate(growth = height_cm - lag(height_cm)) %>%  # find out the diff btw nearby dates
  summarise(has_negative_growth = any(growth < 0, na.rm = TRUE)) %>% 
  filter(has_negative_growth) -> neg_growth_plants
neg_growth_plants

#find out tolerance value
neg_growth_values <- plant_growth %>%
  arrange(PID, survey_date) %>%
  group_by(PID) %>%
  mutate(growth = height_cm - lag(height_cm)) %>%
  ungroup() %>%
  filter(growth < 0)

ggplot(neg_growth_values, aes(x = growth)) +
  geom_histogram(binwidth = 0.5, fill = "red", color = "black") +
  labs(
    title = "Figure A: Distribution of Negative Growth Values",
    x = "Height Decrease (cm)",
    y = "Count"
  ) +
  theme_bw()

#filter out plants with negative growth < -5
plant_growth_cleaned <- plant_growth

repeat {
  plant_growth_cleaned <- plant_growth_cleaned %>%
    arrange(PID, survey_date) %>%
    group_by(PID) %>%
    mutate(growth = height_cm - lag(height_cm)) %>%
    filter(is.na(growth) | growth >= -5) %>%
    select(-growth) %>%
    ungroup()

  check <- plant_growth_cleaned %>%
    arrange(PID, survey_date) %>%
    group_by(PID) %>%
    mutate(growth = height_cm - lag(height_cm)) %>%
    filter(growth < -5)
  
  if (nrow(check) == 0) break
}
```

*Measure Growth via Daily Growth Rate*

```{r}
#define daily growth rate
plant_growth_daily <- plant_growth_cleaned %>%
  arrange(PID, survey_date) %>%
  group_by(PID) %>%
  mutate(
    prev_height = lag(height_cm),
    prev_date = lag(survey_date),
    days_elapsed = as.numeric(survey_date - prev_date),
    daily_growth = (height_cm - prev_height) / days_elapsed
  ) %>%
  ungroup()
```

*Correlate Growth with Solar Radiation*
#Q: How does plant growth correlate with solar radiation?
#Test: Find correlation value and plot correlation
#Figure B: Scatter plot showing the relationship between weekly average solar radiation (SlrW, W/m²) and daily growth rate (cm/day). Each point represents an observation, and the red line indicates the fitted linear regression. 

```{r}
#Align plant growth data to week
plant_weekly <- plant_growth_daily %>%
  filter(!is.na(daily_growth), days_elapsed > 0) %>%
  mutate(week = floor_date(survey_date, "week"))

#Adds `weekly_avg_SlrW` to plant data
plant_with_light <- plant_weekly %>%
  left_join(weekly_light, by = "week")

#Calculate correlation
cor_result <- cor(
  plant_with_light$daily_growth,
  plant_with_light$weekly_avg_SlrW,
use = "complete.obs"
)

print(cor_result)

#Plot correlation
ggplot(plant_with_light, aes(x = weekly_avg_SlrW, y = daily_growth)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Figure B: Correlation between Light and Growth", x = "Weekly Avg Light (SlrW)", y = "Daily Growth (cm/day)") +
  theme_bw()
```

*Process data*

```{r}
#Standardization
plant_with_light$weekly_avg_SlrW2 <- scale(plant_with_light$weekly_avg_SlrW, center = TRUE, scale = TRUE)
mean_light <- mean(plant_with_light$weekly_avg_SlrW, na.rm = TRUE)
sd_light   <- sd(plant_with_light$weekly_avg_SlrW, na.rm = TRUE)

#Change Data type
plant_with_light <- plant_with_light %>%
  mutate(
    parent_pop = factor(parent_pop),
    PID        = factor(PID),
    block      = factor(block)
  )
write.csv(plant_with_light, "Data/plant_with_light", row.names=FALSE)
```

*use mixed-effect model to fit relationship between plant growth and light radiation with population as random effect*
#Q: Does weekly solar radiation positively affect daily growth rate across all populations?
#Test: Fit a mixed-effect model with population as a random slope.
```{r}
growth_light.lmer <- lmer(
  daily_growth ~ weekly_avg_SlrW2 +                     
    (1 + weekly_avg_SlrW2 | parent_pop),                                    
  data = plant_with_light, REML = TRUE
)
summary(growth_light.lmer)
```

*overall effect of solar radiation on plant daily growth*
#Figure C: Relationship between weekly average solar radiation and predicted daily growth rate (cm/day), aggregated across all populations. Each hexagon represents the density of observations. The black line shows the fitted regression slope
.
```{r}
#Unscaling
pred_all <- ggpredict(growth_light.lmer, terms = "weekly_avg_SlrW2") %>%
  as.data.frame() %>%
  mutate(light_orig = x * sd_light + mean_light)

#Plot
p_overall <- ggplot() +
  geom_hex(data = plant_with_light,
           aes(x = weekly_avg_SlrW, y = daily_growth), bins = 35) +
  scale_fill_viridis_c(name = "Count") +
  geom_ribbon(data = pred_all,
              aes(x = light_orig, ymin = conf.low, ymax = conf.high),
              alpha = .22, fill = "grey60") +
  geom_line(data = pred_all,
            aes(x = light_orig, y = predicted), linewidth = 1) +
  labs(title = "Figure C: Effect of Weekly Light on Daily Growth (overall)",
       x = "Weekly Avg Light (W/m²)",
       y = "Predicted Daily Growth (cm/day)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
p_overall
```

*Effect of solar radiation on plant daily growth by population*
#Q: Do different populations vary in their growth response to weekly solar radiation?
#Test: For visualization, scatter plots with fitted regression lines were drawn separately for each population.
#Figure D: Light–growth relationship by population.
#Scatter plots show daily growth rate (cm/day) against weekly average solar radiation (W/m²) for 22 populations. Each panel corresponds to one population, with the blue line indicating the fitted linear trend.

```{r}
p_facet <- ggplot(plant_with_light,
                  aes(weekly_avg_SlrW, daily_growth)) +
  facet_wrap(~ parent_pop, ncol = 6) +
  geom_point(alpha = .15, size = .6, color = "grey35") +
  geom_smooth(method = "lm", se = FALSE, linewidth = .8) +
  labs(title = "Figure D: Light–Growth relationship by population",
       x = "Weekly Avg Light (W/m²)",
       y = "Daily Growth (cm/day)") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey95", color = NA),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())
p_facet
```

*Find slope of plant growth and solar radiation for each population*
#Q: Which populations show significantly positive or negligible slopes in the light–growth relationship?
#Method: Extract slopes and 95% confidence intervals for each population from the mixed-effects model
#Figure E. Population-specific slopes with 95% confidence intervals. 
#Forest plot showing estimated slopes of daily growth rate against weekly average solar radiation for 22 populations. Points represent population-specific slopes (cm/day per W/m²) and horizontal lines show 95% CIs. Several populations (e.g., TM2, CP3, FR, WL2, YO4) exhibit significantly positive slopes with CIs entirely above zero, indicating stronger growth response to light. Others (e.g., CC, IH, DPR) have slopes not significantly different from zero, suggesting little or no detectable response to light.

```{r}
#slope and standard deviation of fixed effects
b_fix  <- fixef(growth_light.lmer)["weekly_avg_SlrW2"]
V_fix  <- vcov(growth_light.lmer)["weekly_avg_SlrW2","weekly_avg_SlrW2"]

#slope and standard deviation of random effects
re <- ranef(growth_light.lmer, condVar = TRUE) 
re_pop <- re$parent_pop
postVar <- attr(re$parent_pop, "postVar")           
sl_col <- which(colnames(re_pop) == "weekly_avg_SlrW2")

#create the tibble of slopes for each population
pop_slope <- tibble(
  parent_pop   = rownames(re_pop),
  rand_slope   = re_pop[ , "weekly_avg_SlrW2"],
  rand_var     = sapply(seq_len(dim(postVar)[3]), function(i) postVar[sl_col, sl_col, i])
) %>%
  mutate(
    slope_SD   = b_fix + rand_slope,                        
    se_SD      = sqrt(V_fix + rand_var),               
    lower_SD   = slope_SD - 1.96*se_SD,
    upper_SD   = slope_SD + 1.96*se_SD
  )

sd_light <- sd(plant_with_light$weekly_avg_SlrW, na.rm = TRUE)

pop_slope <- pop_slope %>%
  mutate(
    slope_per_Wm2 = slope_SD / sd_light,
    lower_per_Wm2 = lower_SD / sd_light,
    upper_per_Wm2 = upper_SD / sd_light
  )%>%
  select(parent_pop, slope_per_Wm2, lower_per_Wm2, upper_per_Wm2)
pop_slope
write.csv(pop_slope, "population_slopes.csv", row.names = FALSE)

#plot
ggplot(pop_slope, aes(x = reorder(parent_pop, slope_per_Wm2), y = slope_per_Wm2)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange(aes(ymin = lower_per_Wm2, ymax = upper_per_Wm2), linewidth = .6) +
  coord_flip() +
  labs(title = "Figure E: Population-specific slopes with 95% CI",
       x = "Population", y = "Slope (cm/day per W/m²)") +
  theme_bw()
```

*height_time plot with solar radiation as color*

```{r}
light_daily <- light_raw %>%
  mutate(date = as.Date(TIMESTAMP)) %>%
  group_by(date) %>%
  summarise(SlrW_mean = mean(SlrW_Avg, na.rm = TRUE), .groups = "drop")

df_plot <- plant_with_light %>%
  filter(!is.na(survey_date), !is.na(height_cm), !is.na(weekly_avg_SlrW))

y_rng     <- range(df_plot$height_cm, na.rm = TRUE)
light_rng <- range(light_daily$SlrW_mean, na.rm = TRUE)

light_daily <- light_daily %>%
  mutate(light_y = scales::rescale(SlrW_mean, to = y_rng, from = light_rng))

plant_with_light%>%
#  group_by(parent_pop)%>%
  ggplot(aes(survey_date, height_cm, colour = weekly_avg_SlrW, group=PID)) +
#  geom_line(data = light_daily,
#            aes(x = date, y = light_y),
#            colour = "black", linewidth = 0.5, alpha= 0.25) +
  facet_wrap(~parent_pop)+
  scale_colour_gradientn(                         # low=blue, high=red
    colours = c("#2c7bb6","#abd9e9","#ffffbf","#fdae61","#d7191c"),
    name = "solar radiation (W/m²)")+
  geom_point(alpha=0.25,size=0.5)+
  geom_line(alpha=0.5) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
   scale_y_continuous(
    name = "Height (cm)")+
#    sec.axis = sec_axis(~ scales::rescale(., from = y_rng, to = light_rng),
#                        name = "Solar Radiation (W/m²)"))+
  labs(title = "Figure F.1: Time-Height",
       x = "Survey Date",
       y = "Height (cm)") +
  theme_bw()
```

```{r}
plant_with_light%>%
  filter(parent_pop=="TM2")%>%
  ggplot(aes(survey_date, height_cm, colour = weekly_avg_SlrW, group=PID)) +
  scale_colour_gradientn(                         # low=blue, high=red
    colours = c("#2c7bb6","#abd9e9","#ffffbf","#fdae61","#d7191c"),
    name = "Solar Radiation (W/m²)")+
  geom_point(alpha=1,size=0.5)+
  geom_line(alpha=1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
   scale_y_continuous(
    name = "Height (cm)")+
  labs(title = "Figure F.2: Time-Height for TM2",
       x = "Survey Date",
       y = "Height (cm)") +
  theme_bw()
```

```{r}
plant_with_light%>%
  filter(parent_pop=="CP3")%>%
  ggplot(aes(survey_date, height_cm, colour = weekly_avg_SlrW, group=PID)) +
  scale_colour_gradientn(                         # low=blue, high=red
    colours = c("#2c7bb6","#abd9e9","#ffffbf","#fdae61","#d7191c"),
    name = "Solar Radiation (W/m²)")+
  geom_point(alpha=1,size=0.5)+
  geom_line(alpha=1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
   scale_y_continuous(
    name = "Height (cm)")+
  labs(title = "Figure F.3: Time-Height for CP3",
       x = "Survey Date",
       y = "Height (cm)") +
  theme_bw()
```

```{r}
plant_with_light%>%
  filter(parent_pop=="FR")%>%
  ggplot(aes(survey_date, height_cm, colour = weekly_avg_SlrW, group=PID)) +
  scale_colour_gradientn(                         # low=blue, high=red
    colours = c("#2c7bb6","#abd9e9","#ffffbf","#fdae61","#d7191c"),
    name = "Solar Radiation (W/m²)")+
  geom_point(alpha=1,size=0.5)+
  geom_line(alpha=1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
   scale_y_continuous(
    name = "Height (cm)")+
  labs(title = "Figure F.4: Time-Height for FR",
       x = "Survey Date",
       y = "Height (cm)") +
  theme_bw()
```

